---
title: "Consolidated Analysis: Performance Metrics (Tutorial)"
author: "Costa, W. G."
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: TRUE
    toc_float: TRUE
    theme: flatly
    highlight: haddock
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = "center",
  out.width = "100%",
  dpi = 300
)
```

# 1\. Introduction and Configuration

This document details the computational pipeline designed to benchmark **Genome-Wide Association Studies (GWAS)** against **Machine Learning (ML)** methods. The process is divided into explicit steps to ensure reproducibility.

The goal is to integrate results from multiple models, standardize their importance scores, and calculate key performance metrics (Detection Power, Precision, FPR, Specificity, and F1 Score) across 10 replicates and 50 cross-validation folds.

### Load Packages

```{r package}
library(tidyverse) # Data manipulation (dplyr) and visualization (ggplot2)
library(scales)    # Axis scaling and formatting
library(patchwork) # Composition of multi-panel plots
library(stringr)   # String manipulation
```

# 2. Data Importation

In this step, we load the raw result files. We distinguish between Machine Learning (ML) and GWAS files and apply column standardization immediately upon reading. This is crucial because different software outputs may use varying column names (e.g., "SNP" vs. "Marker", "p.value" vs. "Importance").

### 2.1. Load Machine Learning Data

Machine Learning models (Random Forest, MARS, etc.) output **Importance Scores** (e.g., Gini Impurity, RSS reduction). Since these scores are on different scales (some range from 0-100, others are unbounded), we must load and normalize them.

**Process:**
1.  **Identify File:** We target the consolidated ML results file generated in the previous script.
2.  **Standardize:** We rename columns to a common format (`variable`, `replicate`, `marker`, `overall`).
3.  **Normalize:** We apply a Min-Max scaling (0 to 10) within each fold/replicate to make scores comparable across different algorithms.

```{r load_ml}
# 1. Define the path to the consolidated ML file
# This file was created in the 'genomic_prediction.Rmd' script
ml_files <- "output/ML_Results_Importance_CONSOLIDATED.csv"

# 2. Reading and Cleaning (Explicit Process)
# We use map_dfr to read the file(s) and apply a consistent cleaning function
ml_data <- map_dfr(ml_files, function(f) {
  
  if(!file.exists(f)) return(data.frame())

  # Read CSV
  df <- read.csv(f)
  colnames(df) <- tolower(colnames(df)) # Standardize names to lowercase to avoid case-sensitivity issues
  
  # Explicitly Select and Rename columns
  # This mapping ensures that 'var', 'Variable', or 'TRAIT' all become 'variable'
  df %>%
    rename(
      variable  = any_of(c("var", "variable")),
      replicate = any_of(c("rep", "replicate")),
      marker    = any_of(c("marker", "marker_idx", "snp")),
      overall   = any_of(c("overall", "importance", "score")),
      fold      = any_of(c("fold")),
      method    = any_of(c("method", "model"))
    ) %>%
    mutate(
      # Ensure numeric types for merging
      variable  = as.numeric(variable),
      # robustly clean marker names (e.g., remove "M" prefix if present)
      marker    = as.numeric(str_replace_all(as.character(marker), "[^0-9]", "")),
      replicate = as.numeric(replicate),
      fold      = as.numeric(fold),
      overall   = as.numeric(overall),
      method    = as.character(method),
      type      = "ML" # Tag to distinguish from GWAS later
    ) %>%
    select(variable, method, marker, replicate, fold, overall, type)
})

# 3. Normalization (Rescale 0-10) within each Fold
# Important: Comparison is only valid if scores share a relative scale.
# We rescale the importance so the maximum importance in any fold is 10.
ml_data <- ml_data %>%
  group_by(variable, method, replicate, fold) %>%
  mutate(
    overall = scales::rescale(overall, to = c(0, 10), finite = FALSE)) %>%
  ungroup()

head(ml_data)
```

### 2.2. Load GWAS Data

GWAS models (GLM, FarmCPU, Blink) output **P-values**. These represent the statistical probability of the association being random.

**Process:**

1.  **Identify File:** We target the consolidated GWAS results file generated in the GAPIT script.
2.  **Standardize:** Similar to ML, we standardize column names.
3.  **No Normalization:** Unlike ML scores, P-values (transformed to $-log_{10}(P)$) have a standard statistical interpretation and do not require rescaling.

<!-- end list -->

```{r load_gwas}
# Path to the already unified file from 'models_gapit_gwas.Rmd'
gwas_file <- "output/gwas_cv/GWAS_Results_Importance_CONSOLIDATED.csv"

if (file.exists(gwas_file)) {
  
  # 1. Read single file
  df_raw <- read.csv(gwas_file)
  
  # 2. Standardize column names (to lowercase)
  colnames(df_raw) <- tolower(colnames(df_raw))
  
  # 3. Cleaning and Selection
  gwas_data <- df_raw %>%
    rename(
      # Map file columns (Variable, Model...) to standard pattern (variable, method...)
      variable  = any_of(c("variable", "var")),
      replicate = any_of(c("replicate", "rep")),
      marker    = any_of(c("marker", "snp")),
      overall   = any_of(c("overall", "p.value", "score")), 
      fold      = any_of(c("fold")),
      method    = any_of(c("model", "method"))
    ) %>%
    mutate(
      # Robust type conversion
      variable  = as.numeric(variable),
      # Remove non-numeric characters from marker (e.g., "V1" -> 1)
      marker    = as.numeric(str_replace_all(as.character(marker), "[^0-9]", "")),
      replicate = as.numeric(replicate),
      fold      = as.numeric(fold),
      overall   = as.numeric(overall),
      method    = as.character(method),
      type      = "GWAS" # Tag to differentiate later
    ) %>%
    # Safety filter to remove corrupted rows (NAs)
    filter(!is.na(marker), !is.na(variable)) %>%
    select(variable, method, marker, replicate, fold, overall, type)
  
  cat("GWAS data loaded successfully.\n")
  cat("Total records:", nrow(gwas_data), "\n")
  
} else {
  warning("File 'GWAS_Results_Importance_CONSOLIDATED.csv' not found in output/gwas_cv/.")
  gwas_data <- data.frame()
}
```

### 2.3. Unification

Finally, we stack the two datasets vertically. This creates a master dataset containing every tested marker, for every method, variable, and replicate, ready for the comparative analysis.

```{r unify}
# Combine ML and GWAS dataframes
all_data <- bind_rows(ml_data, gwas_data)
cat("Total consolidated records:", nrow(all_data), "\n")
```

# 3. Ground Truth Definition (True QTLs)

To evaluate the methods, we must define the locations of the true simulated QTLs. We load the base simulation file and expand it to cover all 18 variables, representing different heritability scenarios ($h^2 = 0.3, 0.5, 0.8$).

We create two reference structures:
1.  **Flat Map (`region_map`):** Used for calculating **Precision** and **Specificity**. Overlapping windows are merged to prevent double-counting markers as "False Positives".
2.  **ID Reference (`qtl_reference`):** Used for calculating **Detection Power**. Preserves unique QTL IDs to determine how many distinct genetic signals were successfully detected.

```{r ground_truth}
# Load original file (Contains variables 1 to 6)
if(file.exists("data/snpsOfInterest.RData")) {
  load("data/snpsOfInterest.RData")
} else {
  stop("File 'snpsOfInterest.RData' not found in data/ folder.")
}

# Expansion to include different heritability scenarios
# Variables 1-6  (h2=0.3)
# Variables 7-12 (h2=0.5) - Shifted Copy (+6)
# Variables 13-18 (h2=0.8) - Shifted Copy (+12)
snpsOfInterest <- snpsOfInterest %>%
  bind_rows(snpsOfInterest %>% mutate(variable = variable + 6)) %>% 
  bind_rows(snpsOfInterest %>% mutate(variable = variable + 12))

# Window Parameters
distance <- 5 
total_markers_genotype <- 4010 

# --- A. Flat Map (For Precision/FP/TN) ---
# We use distinct() to merge overlapping windows into a single boolean "target" region
region_map <- snpsOfInterest %>%
  rowwise() %>%
  mutate(marker = list(seq(marker - distance, marker + distance))) %>%
  ungroup() %>%
  unnest(cols = marker) %>%
  select(variable, marker) %>% 
  distinct() %>% # REMOVE DUPLICATE markers (merges overlaps)
  mutate(
    marker = as.numeric(marker),
    is_in_region = TRUE
  )

# --- B. Reference with ID (For Detection Power) ---
# We keep the QTL_ID to count how many "unique QTLs" were found (Hit Rate)
qtl_reference <- snpsOfInterest %>%
  mutate(QTL_ID = paste0("QTL_", variable, "_", marker)) %>%
  rowwise() %>%
  mutate(marker = list(seq(marker - distance, marker + distance))) %>%
  ungroup() %>%
  unnest(cols = marker) %>%
  select(variable, marker, QTL_ID)

# --- C. Denominators (For Rates) ---
truth_metadata <- snpsOfInterest %>%
  group_by(variable) %>%
  summarise(total_QTLs_real = n_distinct(marker), .groups = "drop")

markers_in_target <- region_map %>%
  group_by(variable) %>%
  summarise(total_target_markers = n_distinct(marker))

truth_metadata <- left_join(truth_metadata, markers_in_target, by = "variable") %>%
  mutate(total_noise_markers = total_markers_genotype - total_target_markers)
```

# 4. Marker Classification

In this step, we apply specific selection thresholds to classify markers as "Significant" (Selected) or "Not Significant".

* **GWAS:** We apply the rigorous **Bonferroni correction** ($-\log_{10}(0.05/M)$) to control the family-wise error rate.
* **Machine Learning:** We use an **empirical threshold** based on the mean importance score within each replicate. Markers with importance above the mean are selected.

Finally, we cross-reference the selected markers with the `region_map` to categorize them into the Confusion Matrix classes: True Positives (TP), False Positives (FP), False Negatives (FN), and True Negatives (TN).

```{r classification}
# Identify GWAS methods from the data
gwas_methods <- unique(gwas_data$method)

# Calculate Thresholds and Selection Status
classified_data <- all_data %>%
  group_by(method, variable, replicate, fold) %>%
  mutate(
    threshold_ml = mean(overall, na.rm=TRUE),
    threshold_gwas = -log10(0.05/total_markers_genotype), 
    
    is_selected = case_when(
      # GWAS Criteria: > Bonferroni
      method %in% gwas_methods & overall > threshold_gwas ~ TRUE,
      # ML Criteria: > Mean Importance
      !(method %in% gwas_methods) & overall > threshold_ml ~ TRUE,
      TRUE ~ FALSE
    )
  ) %>%
  ungroup() %>%
  # Join with Flat Map (Left Join to retain all tested markers)
  left_join(region_map, by = c("variable", "marker")) %>%
  mutate(
    # Markers not found in region_map are "Noise" (FALSE)
    is_in_region = replace_na(is_in_region, FALSE),
    
    # Final Classification (Confusion Matrix)
    classification = case_when(
      is_selected & is_in_region  ~ "TP", # True Positive (Hit)
      is_selected & !is_in_region ~ "FP", # False Positive (Noise Selection)
      !is_selected & is_in_region ~ "FN", # False Negative (Missed QTL)
      !is_selected & !is_in_region ~ "TN" # True Negative (Correct Rejection)
    )
  )
```

# 5. Calculation of Metrics by Replicate

In this step, we calculate the raw counts of "hits" and "misses" for each cross-validation fold.

### 5.1. QTL Hits (For Detection Power and Precision)

We use the `qtl_reference` table (which contains unique QTL IDs) to determine how many distinct QTLs were "tagged" by the selected markers. 

```{r calc_hits}
hits_calc <- classified_data %>%
  filter(is_selected) %>% 
  # relationship = "many-to-many" is required because QTL windows may overlap
  inner_join(qtl_reference, by = c("variable", "marker"), relationship = "many-to-many") %>%
  group_by(variable, method, replicate, fold) %>%
  summarise(acerto_qtl = n_distinct(QTL_ID), .groups = "drop")
```

### 5.2. Error Statistics (FP) and Rejections (TN)

We summarize the marker-level classification to count False Positives (Selection of Noise) and True Negatives (Correct Rejection of Noise).

[Image of confusion matrix precision recall]

```{r calc_stats}
marker_stats <- classified_data %>%
  group_by(variable, method, replicate, fold) %>%
  summarise(
    # FP_markers = Error (Selected noise marker)
    FP_markers = sum(classification == "FP"),
    # TN_markers = Rejection Hit (Correctly ignored noise marker)
    TN_markers = sum(classification == "TN"),
    .groups = "drop"
  )
```

# 6. Consolidation and Calculation of Means

In this step, we merge all classification data and calculate the performance metrics for each replicate. Subsequently, we aggregate the results to obtain the Mean ($\mu$) and Standard Deviation ($\sigma$) across the 10 replicates.

```{r consolidation}
# Create skeleton to ensure methods with zero selections are included in the stats
skeleton <- classified_data %>% distinct(variable, method, replicate, fold)

# Merge all components
coinc_raw <- skeleton %>%
  left_join(truth_metadata, by = "variable") %>%
  left_join(hits_calc, by = c("variable", "method", "replicate", "fold")) %>%
  left_join(marker_stats, by = c("variable", "method", "replicate", "fold")) %>%
  mutate(
    # Replace NAs with 0 (for methods that selected nothing)
    acerto_qtl = replace_na(acerto_qtl, 0),
    FP_markers = replace_na(FP_markers, 0), 
    TN_markers = replace_na(TN_markers, 0),
    
    # --- METRIC FORMULAS ---
    
    # 1. Detection Power (DP)
    # (Unique QTLs Found / Total Real QTLs) * 100
    DP = (acerto_qtl / total_QTLs_real) * 100,
    
    # 2. Precision
    # (True Hits / Total Selected Signals) * 100
    Precision = ifelse((acerto_qtl + FP_markers) == 0, 0, 
                       (acerto_qtl / (acerto_qtl + FP_markers)) * 100),
    
    # 3. False Positive Rate (FP Rate)
    # (False Positives / Total Noise Markers) * 100
    FP_Rate = ifelse(total_noise_markers == 0, 0, 
                     (FP_markers / total_noise_markers) * 100),
    
    # 4. Specificity
    # (True Negatives / Total Noise Markers) * 100
    Specificity = ifelse(total_noise_markers == 0, 0,
                         ((total_noise_markers - FP_markers) / total_noise_markers) * 100),
    
    # 5. F1 Score (Harmonic Mean of DP and Precision)
    F1_Score = ifelse((DP + Precision) == 0, 0, 
                      2 * ((DP * Precision) / (DP + Precision)))
  )

# Calculate Means and Standard Deviations across replicates
coinc_summary <- coinc_raw %>%
  group_by(variable, method) %>%
  summarise(
    DP_mean = mean(DP, na.rm=TRUE),
    DP_sd = sd(DP, na.rm=TRUE),
    
    Precision_mean = mean(Precision, na.rm=TRUE),
    Precision_sd = sd(Precision, na.rm=TRUE),
    
    FP_mean = mean(FP_Rate, na.rm=TRUE),
    FP_sd = sd(FP_Rate, na.rm=TRUE),
    
    Specificity_mean = mean(Specificity, na.rm=TRUE),
    Specificity_sd = sd(Specificity, na.rm=TRUE),
    
    F1_Score_mean = mean(F1_Score, na.rm=TRUE),
    F1_Score_sd = sd(F1_Score, na.rm=TRUE),
    
    .groups = "drop"
  )

# Add Metadata and Classification (MM vs ML)
coinc_plot <- coinc_summary %>%
  mutate(
    # Classify Heritability
    herd = case_when(
      variable %in% 1:6  ~ "h2 = 0.3",
      variable %in% 7:12 ~ "h2 = 0.5",
      variable %in% 13:18 ~ "h2 = 0.8",
      TRUE ~ "Unknown"
    ),
    # Classify Number of Genes
    ngenes = case_when(
      variable %in% c(1, 7, 13) ~ 8,
      variable %in% c(2, 8, 14) ~ 40,
      variable %in% c(3, 9, 15) ~ 80,
      variable %in% c(4, 10, 16) ~ 120,
      variable %in% c(5, 11, 17) ~ 240,
      variable %in% c(6, 12, 18) ~ 480,
      TRUE ~ 999
    ),
    # CORRECTION: Classify GAPIT models as "MM" (Mixed/Statistical Models)
    methodology = ifelse(method %in% gwas_methods, "MM", "ML")
  )

# Pre-format plot labels (Mean ± SD)
coinc_plot <- coinc_plot %>%
  mutate(
    label_dp   = paste0(round(DP_mean, 1), " ± ", round(DP_sd, 1)),
    label_prec = paste0(round(Precision_mean, 1), " ± ", round(Precision_sd, 1)),
    label_fp   = paste0(round(FP_mean, 2), " ± ", round(FP_sd, 2)),
    label_spec = paste0(round(Specificity_mean, 1), " ± ", round(Specificity_sd, 1)),
    label_f1   = paste0(round(F1_Score_mean, 1), " ± ", round(F1_Score_sd, 1)),
    
    # Factor Ordering for Plotting
    method = factor(method, levels = rev(sort(unique(method)))),
    ngenes_f = factor(ngenes, levels = sort(unique(ngenes)))
  )
```

```{r filter_variables, include=FALSE}
coinc_plot <- coinc_plot %>%
  filter(variable %in% c(7:11, 13:17)) # Ensure only valid variables are included
```

# 7. Visualization (Heatmaps)

Below, we generate comparative heatmaps between **Mixed Models (MM)** and **Machine Learning (ML)**.

### 7.1. Detection Power (DP)

```{r plot_dp, fig.width=14, fig.height=8}
# Plot MM (Statistical Models)
p1 <- coinc_plot %>% filter(methodology == "MM") %>%
  ggplot(aes(x = ngenes_f, y = method, fill = DP_mean)) +
  geom_tile(color = "black") +
  geom_text(aes(label = label_dp), color = "black", size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", midpoint = 50, limits = c(0, 100), name = "DP (%)") +
  facet_grid(herd ~ ., scales = "free", space = "free") +
  labs(title = "Mixed Models (MM)", x = NULL, y = NULL) +
  theme_bw() + 
  theme(legend.position = "none", axis.text.y = element_text(size = 10))

# Plot ML (Machine Learning)
p2 <- coinc_plot %>% filter(methodology == "ML") %>%
  ggplot(aes(x = ngenes_f, y = method, fill = DP_mean)) +
  geom_tile(color = "black") +
  geom_text(aes(label = label_dp), color = "black", size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", midpoint = 50, limits = c(0, 100), name = "DP (%)") +
  facet_grid(herd ~ ., scales = "free", space = "free") +
  labs(title = "Machine Learning (ML)", x = NULL, y = NULL) +
  theme_bw() + 
  theme(axis.text.y = element_text(size = 10), axis.ticks.y = element_line())

final_dp <- (p1 | p2) + plot_annotation(title = "Detection Power", theme = theme(plot.title = element_text(hjust = 0.5, face="bold", size=16)))

print(final_dp)
ggsave("output/DP_heatmap.tiff", final_dp, width = 14, height = 8, dpi = 300)
```

### 7.2. Precision

```{r plot_prec, fig.width=14, fig.height=8}
p1 <- coinc_plot %>% filter(methodology == "MM") %>%
  ggplot(aes(x = ngenes_f, y = method, fill = Precision_mean)) +
  geom_tile(color = "black") +
  geom_text(aes(label = label_prec), color = "black", size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", midpoint = 50, limits = c(0, 100), name = "Precision (%)") +
  facet_grid(herd ~ ., scales = "free", space = "free") +
  labs(title = "Mixed Models (MM)", x = NULL, y = NULL) +
  theme_bw() + theme(legend.position = "none", axis.text.y = element_text(size = 10))

p2 <- coinc_plot %>% filter(methodology == "ML") %>%
  ggplot(aes(x = ngenes_f, y = method, fill = Precision_mean)) +
  geom_tile(color = "black") +
  geom_text(aes(label = label_prec), color = "black", size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", midpoint = 50, limits = c(0, 100), name = "Precision (%)") +
  facet_grid(herd ~ ., scales = "free", space = "free") +
  labs(title = "Machine Learning (ML)", x = NULL, y = NULL) +
  theme_bw() + theme(axis.text.y = element_text(size = 10), axis.ticks.y = element_line())

final_prec <- (p1 | p2) + plot_annotation(title = "Precision", theme = theme(plot.title = element_text(hjust = 0.5, face="bold", size=16)))

print(final_prec)
ggsave("output/Precision_heatmap.tiff", final_prec, width = 14, height = 8, dpi = 300)
```

### 7.3. False Positive Rate (FPR)

```{r plot_fp, fig.width=14, fig.height=8}
p1 <- coinc_plot %>% filter(methodology == "MM") %>%
  ggplot(aes(x = ngenes_f, y = method, fill = FP_mean)) +
  geom_tile(color = "black") +
  geom_text(aes(label = label_fp), color = "black", size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", midpoint = 50, limits = c(0, 100), name = "FPR (%)") +
  facet_grid(herd ~ ., scales = "free", space = "free") +
  labs(title = "Mixed Models (MM)", x = NULL, y = NULL) +
  theme_bw() + theme(legend.position = "none", axis.text.y = element_text(size = 10))

p2 <- coinc_plot %>% filter(methodology == "ML") %>%
  ggplot(aes(x = ngenes_f, y = method, fill = FP_mean)) +
  geom_tile(color = "black") +
  geom_text(aes(label = label_fp), color = "black", size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", midpoint = 50, limits = c(0, 100), name = "FPR (%)") +
  facet_grid(herd ~ ., scales = "free", space = "free") +
  labs(title = "Machine Learning (ML)", x = NULL, y = NULL) +
  theme_bw() + theme(axis.text.y = element_text(size = 10), axis.ticks.y = element_line())

final_fp <- (p1 | p2) + plot_annotation(title = "False Positive Rate (FPR)", theme = theme(plot.title = element_text(hjust = 0.5, face="bold", size=16)))

print(final_fp)
ggsave("output/FP_heatmap.tiff", final_fp, width = 14, height = 8, dpi = 300)
```

### 7.4. Specificity

```{r plot_spec, fig.width=14, fig.height=8}
p1 <- coinc_plot %>% filter(methodology == "MM") %>%
  ggplot(aes(x = ngenes_f, y = method, fill = Specificity_mean)) +
  geom_tile(color = "black") +
  geom_text(aes(label = label_spec), color = "black", size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", midpoint = 50, limits = c(0, 100), name = "Specificity (%)") +
  facet_grid(herd ~ ., scales = "free", space = "free") +
  labs(title = "Mixed Models (MM)", x = NULL, y = NULL) +
  theme_bw() + theme(legend.position = "none", axis.text.y = element_text(size = 10))

p2 <- coinc_plot %>% filter(methodology == "ML") %>%
  ggplot(aes(x = ngenes_f, y = method, fill = Specificity_mean)) +
  geom_tile(color = "black") +
  geom_text(aes(label = label_spec), color = "black", size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", midpoint = 50, limits = c(0, 100), name = "Specificity (%)") +
  facet_grid(herd ~ ., scales = "free", space = "free") +
  labs(title = "Machine Learning (ML)", x = NULL, y = NULL) +
  theme_bw() + theme(axis.text.y = element_text(size = 10), axis.ticks.y = element_line())

final_spec <- (p1 | p2) + plot_annotation(title = "Specificity", theme = theme(plot.title = element_text(hjust = 0.5, face="bold", size=16)))

print(final_spec)
ggsave("output/Specificity_heatmap.tiff", final_spec, width = 14, height = 8, dpi = 300)
```

### 7.5. F1 Score

```{r plot_f1, fig.width=14, fig.height=8}
p1 <- coinc_plot %>% filter(methodology == "MM") %>%
  ggplot(aes(x = ngenes_f, y = method, fill = F1_Score_mean)) +
  geom_tile(color = "black") +
  geom_text(aes(label = label_f1), color = "black", size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", midpoint = 50, limits = c(0, 100), name = "F1 Score") +
  facet_grid(herd ~ ., scales = "free", space = "free") +
  labs(title = "Mixed Models (MM)", x = NULL, y = NULL) +
  theme_bw() + theme(legend.position = "none", axis.text.y = element_text(size = 10))

p2 <- coinc_plot %>% filter(methodology == "ML") %>%
  ggplot(aes(x = ngenes_f, y = method, fill = F1_Score_mean)) +
  geom_tile(color = "black") +
  geom_text(aes(label = label_f1), color = "black", size = 2.8) +
  scale_fill_gradient2(low = "white", mid = "yellow", high = "red", midpoint = 50, limits = c(0, 100), name = "F1 Score") +
  facet_grid(herd ~ ., scales = "free", space = "free") +
  labs(title = "Machine Learning (ML)", x = NULL, y = NULL) +
  theme_bw() + theme(axis.text.y = element_text(size = 10), axis.ticks.y = element_line())

final_f1 <- (p1 | p2) + plot_annotation(title = "F1 Score (Harmonic Mean)", theme = theme(plot.title = element_text(hjust = 0.5, face="bold", size=16)))

print(final_f1)
ggsave("output/F1_Score_heatmap.tiff", final_f1, width = 14, height = 8, dpi = 300)
```

# 8. Tabela Detalhada: Acertos e Erros

Abaixo apresentamos a média e o desvio padrão de:
* **Acertos (TP):** Quantidade de QTLs únicos encontrados.
* **Erros (FP):** Quantidade de marcadores de ruído selecionados incorretamente.

```{r detailed_table}
# 1. Preparar os dados brutos
# Unimos as contagens de Hits (QTLs) e Stats (FP/TN)
raw_counts <- hits_calc %>%
  full_join(marker_stats, by = c("variable", "method", "replicate", "fold")) %>%
  mutate(
    acerto_qtl = replace_na(acerto_qtl, 0),
    FP_markers = replace_na(FP_markers, 0)
  )

# 2. Calcular Média e Desvio Padrão por Variável e Método
table_summary <- raw_counts %>%
  group_by(variable, method) %>%
  summarise(
    # Acertos (TP)
    TP_Mean = mean(acerto_qtl),
    TP_SD   = sd(acerto_qtl),
    
    # Erros (FP)
    FP_Mean = mean(FP_markers),
    FP_SD   = sd(FP_markers),
    
    .groups = "drop"
  )

# 3. Adicionar Metadados para facilitar a leitura
table_display <- table_summary %>%
  mutate(
    # Formatação "Média ± SD"
    Hits_Format = paste0(round(TP_Mean, 2), " ± ", round(TP_SD, 2)),
    Errors_Format = paste0(round(FP_Mean, 2), " ± ", round(FP_SD, 2)),
    
    # Informação de Herdabilidade (Opcional, ajuda na filtragem)
    Heritability = case_when(
      variable %in% 1:6  ~ "0.3",
      variable %in% 7:12 ~ "0.5",
      variable %in% 13:18 ~ "0.8",
      TRUE ~ "-"
    ),
    Methodology = ifelse(method %in% gwas_methods, "MM", "ML")
  ) %>%
  select(variable, Heritability, method, Methodology, Hits_Format, Errors_Format) %>%
  arrange(variable, desc(Methodology), method) %>% 
  filter(variable %in% c(7:11, 13:17)) # Filtrar apenas variáveis válidas

# 4. Gerar Tabela Interativa
# Se não tiver o pacote DT instalado: install.packages("DT")
library(DT)

datatable(table_display, 
          rownames = FALSE,
          filter = 'top',
          options = list(pageLength = 15, autoWidth = TRUE),
          colnames = c("Var", "h2", "Method", "Type", "Hits (Mean ± SD)", "Errors (Mean ± SD)"))
```
## General Conclusion

The present analysis allows for the following conclusions:

  * **Sensitivity vs. Precision:** Non-parametric regression methods (such as MARS) demonstrated a superior balance between Detection Power and Precision when compared to classic Mixed Models and simple Decision Trees, particularly in complex scenarios with high polygenicity.
  * **Stability (Standard Deviation):** The inclusion of standard deviation in the metrics reveals that while some ML methods achieve high mean performance, their variability across replicates can compromise reproducibility, a critical factor for genomic selection.
  * **False Positives:** Machine Learning methods tend to have a false positive rate controlled by the empirical mean threshold, but often higher than the rigorous Bonferroni criterion used in Mixed Models, highlighting the need for biological validation of candidates selected by ML.

