---
title: "Genomic Prediction & Marker Importance (Machine Learning)"
author: "Costa, W. G."
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: TRUE
    toc_float: TRUE
    theme: flatly
    highlight: haddock
editor_options:
  chunk_output_type: console
---

# Machine Learning for QTL Detection

This script performs genomic prediction using various Machine Learning algorithms. The primary goal is to extract **Marker Importance Scores** to evaluate their ability to detect QTLs compared to GWAS methods.

**Models:** Random Forest (RF), Bagging, Boosting (GBM), Decision Tree (DT), MARS.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = "center"
)
```

## 1\. Setup and Packages

We load the necessary libraries for data manipulation, parallel processing, and specific Machine Learning algorithms.

```{r packages, message=FALSE}
# Data Manipulation & Parallelism
library(tidyverse)  # Data handling
library(gtools)     # Natural sorting of files
library(doParallel) # Parallel backend
library(foreach)    # Parallel loops

# Machine Learning Models - Install if missing
if(!require(randomForest)) install.packages("randomForest") # RF & Bagging
if(!require(gbm)) install.packages("gbm")                   # Boosting (Gradient Boosting Machine)
if(!require(earth)) install.packages("earth")               # MARS (Multivariate Adaptive Regression Splines)
if(!require(rpart)) install.packages("rpart")               # Decision Trees

# Load ML Libraries
library(randomForest)
library(gbm)
library(earth)
library(rpart)
```

## 2. Data Preparation

We identify and list the partitioned files located in the `PHEN_T` (Phenotype) and `GEN_T` (Genotype) directories. We use natural sorting to ensure correct file ordering and perform integrity checks to guarantee that phenotype and genotype files are correctly paired.

```{r load_data}
# 1. Define File Paths
pathFenT <- file.path("data/PHEN_T")
pathGenT <- file.path("data/GEN_T")

# 2. File Listing (Natural Numeric Sort)
# 'mixedsort' ensures that "file10" comes after "file2", rather than after "file1"
filesFenT <- gtools::mixedsort(list.files(path = pathFenT, pattern = "\\.txt$", full.names = TRUE))
filesGenT <- gtools::mixedsort(list.files(path = pathGenT, pattern = "\\.txt$", full.names = TRUE))

# Integrity Check
if(length(filesFenT) == 0 || length(filesGenT) == 0) {
  stop("CRITICAL ERROR: No phenotype or genotype files found in data/ folders.")
}

if(length(filesFenT) != length(filesGenT)) {
  warning("WARNING: Mismatch in number of phenotype and genotype files. Check pairing.")
}

cat("Files detected: ", length(filesFenT), "pairs (Phenotype/Genotype).\n")
```

## 3. Global Parameters and Cluster Setup

We define the experimental design (10 replicates, 5 folds) and the list of Machine Learning models to be evaluated. We also configure the parallel backend to utilize available processor cores efficiently.

```{r grid_setup}
# --- Experimental Parameters ---
nfolds <- 5
n_reps <- 10
total_traits <- 18
target_vars <- 1:total_traits 

# Machine Learning Models
# RF = Random Forest, BA = Bagging, BO = Boosting, DT = Decision Tree, MARS = Earth
ml_models <- c("RF", "BA", "BO", "DT", "MARS")

# Output Directories
base_out_dir <- "output/ml_cv"
temp_run_dir <- file.path(base_out_dir, "temp_runs") 
if (!dir.exists(temp_run_dir)) dir.create(temp_run_dir, recursive = TRUE)

# --- Cluster Configuration ---
# Detect physical cores to avoid overloading (logical=FALSE is safer for RAM-heavy ML)
n_cores_total <- parallel::detectCores()
n_cores_use <- max(1, n_cores_total - 1)

cat("Cluster Configuration: Using", n_cores_use, "cores.\n")

# Restart cluster if it already exists to avoid conflicts
if(exists("cl")) try(stopCluster(cl), silent=TRUE)
cl <- makeCluster(n_cores_use)
registerDoParallel(cl)
```

## 4\. Machine Learning Execution

### 4.1. Multivariate Adaptive Regression Splines (MARS)

MARS is a non-parametric regression technique that models non-linear relationships by fitting piecewise linear segments (hinge functions). It automatically performs feature selection by "pruning" terms that do not contribute to model stability.

```{r mars_run, eval=FALSE}
# 1. Setup
grau_mars <- 1:3 
grid_mars <- expand.grid(Rep = 1:n_reps, Variable = target_vars, Degree = grau_mars)
out_mars <- paste0(base_out_dir,"/mars")
if (!dir.exists(out_mars)) dir.create(out_mars, recursive = TRUE)

cat("Iniciando MARS. Total de tarefas:", nrow(grid_mars) * nfolds, "\n")

set.seed(123) 

# 2. Execução
mars_list <- foreach(task_i = 1:nrow(grid_mars), 
                     .packages = c('earth', 'caret', 'tidyverse'),
                     .export = c("filesFenT", "filesGenT", "filesFenV", "filesGenV", "nfolds")) %dopar% {
  
  curr_rep <- grid_mars$Rep[task_i]
  curr_var <- grid_mars$Variable[task_i]
  curr_deg <- grid_mars$Degree[task_i]
  
  # Definir nome composto do modelo
  model_name_compound <- paste("MARS", curr_deg)
  
  # Acumuladores locais (opcional, pois salvamos CSV)
  fold_metrics_accum <- data.frame()
  fold_importance_accum <- data.frame()
  
  for (fold_k in 1:nfolds) {
    
    # A. Lógica de Reciclagem de Arquivos
    raw_idx <- (curr_rep - 1) * nfolds + fold_k
    file_idx <- ((raw_idx - 1) %% length(filesFenT)) + 1
    
    # Checkpoint: Pula se já existir o arquivo de importância
    outfile_imp <- sprintf("%s/IMP_Rep%d_Var%d_Deg%d_Fold%d.csv", out_mars, curr_rep, curr_var, curr_deg, fold_k)
    if(file.exists(outfile_imp)) next
    
    tryCatch({
      # B. Leitura dos Dados (Sem renomear colunas)
      # O R usará o padrão: V1, V2, V3...
      dadosTy <- read.table(filesFenT[file_idx], header = FALSE)
      dadosTx <- read.table(filesGenT[file_idx], header = FALSE)
      dadosVy <- read.table(filesFenV[file_idx], header = FALSE)
      dadosVx <- read.table(filesGenV[file_idx], header = FALSE)
      
      start_time <- Sys.time()
      
      # C. Modelo
      mars_model <- earth(x = dadosTx, y = dadosTy[, curr_var], degree = curr_deg)
      
      exec_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
      
      # --- D. CÁLCULO DE MÉTRICAS ---
      ypred <- predict(mars_model, dadosVx)[,1] # [,1] garante vetor
      ypred_train <- mars_model$fitted.values[,1]
      
      # Validação
      if (sd(ypred) == 0) { 
        r2_val <- 0 
      } else { 
        rv <- cor(dadosVy[, curr_var], ypred)
        r2_val <- as.numeric(rv^2)
      }
      rmse_val <- as.numeric(sqrt(mean((dadosVy[, curr_var] - ypred)^2)))
      
      # Treinamento
      rt <- cor(dadosTy[, curr_var], ypred_train)
      r2_train <- as.numeric(rt^2)
      rmse_train <- as.numeric(sqrt(mean((dadosTy[, curr_var] - ypred_train)^2)))
      
      # Salvar Métricas
      curr_metrics <- data.frame(
        Model = model_name_compound, 
        Replicate = curr_rep, 
        Variable = curr_var, 
        Degree = curr_deg, 
        Fold = fold_k, 
        R2_Train = r2_train, 
        R2_Val = r2_val, 
        RMSE_Train = rmse_train, 
        RMSE_Val = rmse_val, 
        Time_Sec = exec_time
      )
      
      write.csv(curr_metrics, sprintf("%s/METRICS_Rep%d_Var%d_Deg%d_Fold%d.csv", out_mars, curr_rep, curr_var, curr_deg, fold_k), row.names = FALSE)
      
      # --- E. Importância (Código Solicitado) ---
      
      imp <- tryCatch({ data.frame(varImp(mars_model, value = "rss")) }, error=function(e) NULL)
      
      if (!is.null(imp)) {
         # --- PADRONIZAÇÃO DE COLUNAS (IMPORTÂNCIA) ---
         imp$Marker <- rownames(imp)
         # Se quiser remover o "V" padrão do R (V1 -> 1), descomente abaixo:
         # imp$Marker <- gsub("V", "", imp$Marker) 
         
         colnames(imp)[1] <- "Overall"
         
         imp_final <- data.frame(
           Model = model_name_compound,
           Replicate = curr_rep,
           Variable = curr_var,
           Degree = curr_deg,
           Fold = fold_k,
           Marker = imp$Marker,
           Overall = imp$Overall # Usando "rss" como medida de importância 
         )
         
         write.csv(imp_final, outfile_imp, row.names = FALSE)
      } else {
         # Salva placeholder vazio se nenhuma variável for selecionada
         dummy <- data.frame(Model=model_name_compound, Overall=0)
         write.csv(dummy, outfile_imp, row.names=FALSE)
      }
      
    }, error = function(e) {
       writeLines(paste("Erro no MARS:", e$message), file.path(out_mars, "error_log.txt"))
    })
    
    # Limpeza de memória
    rm(mars_model, dadosTx, dadosVx, dadosTy, dadosVy); gc()
  }
  return(NULL)
}

# 3. Consolidação Final (Executada após o loop paralelo)
cat("Consolidando resultados do MARS...\n")

# Listar arquivos gerados
metric_files <- list.files(out_mars, pattern = "METRICS_.*\\.csv", full.names = TRUE)
imp_files    <- list.files(out_mars, pattern = "IMP_.*\\.csv", full.names = TRUE)

if(length(metric_files) > 0) {
  mars_metrics <- map_dfr(metric_files, read.csv)
  write.csv(mars_metrics, paste0(out_mars, "/MARS_Results_Metrics_CONSOLIDATED.csv"), row.names = FALSE)
}

if(length(imp_files) > 0) {
  mars_imp <- map_dfr(imp_files, read.csv)
  write.csv(mars_imp, paste0(out_mars, "/MARS_Results_Importance_CONSOLIDATED.csv"), row.names = FALSE)
}

cat("MARS finalizado.\n")
```

### 4.2. Decision Tree (DT)

Decision Trees partition the genetic data into recursive binary splits. The importance of a marker is calculated based on the improvement in the splitting criterion (e.g., ANOVA for regression) summed over all nodes where that marker was used.

```{r dt_run, eval=FALSE}
# 1. Setup
grid_dt <- expand.grid(Rep = 1:n_reps, Variable = target_vars, stringsAsFactors = FALSE)
out_dt <- paste0(base_out_dir,"/dt")
if (!dir.exists(out_dt)) dir.create(out_dt, recursive = TRUE)

cat("Iniciando DT. Total de tarefas:", nrow(grid_dt) * nfolds, "\n")

set.seed(123)

# 2. Execução
dt_list <- foreach(task_i = 1:nrow(grid_dt), 
                   .packages = c('rpart', 'caret', 'tidyverse'),
                   .export = c("filesFenT", "filesGenT", "filesFenV", "filesGenV", "nfolds")) %dopar% {
  
  curr_rep <- grid_dt$Rep[task_i]
  curr_var <- grid_dt$Variable[task_i]
  
  # Acumuladores locais (para manter compatibilidade com o snippet, embora salvemos em disco)
  fold_metrics_accum <- data.frame()
  fold_importance_accum <- data.frame()
  
  for (fold_k in 1:nfolds) {
    
    # A. Lógica de Reciclagem de Arquivos
    raw_idx <- (curr_rep - 1) * nfolds + fold_k
    file_idx <- ((raw_idx - 1) %% length(filesFenT)) + 1
    
    # Checkpoint: Pula se já existir o arquivo de importância
    outfile_imp <- sprintf("%s/IMP_Rep%d_Var%d_Fold%d.csv", out_dt, curr_rep, curr_var, fold_k)
    if(file.exists(outfile_imp)) next
    
    tryCatch({
      # B. Leitura dos Dados
      # O R usará o padrão: V1, V2, V3... para o Genótipo
      dadosTy <- read.table(filesFenT[file_idx], header = FALSE)
      dadosTx <- read.table(filesGenT[file_idx], header = FALSE)
      dadosVy <- read.table(filesFenV[file_idx], header = FALSE)
      dadosVx <- read.table(filesGenV[file_idx], header = FALSE)
      
      start_time <- Sys.time()
      
      # C. Modelo (rpart requer dataframe único ou formula)
      # Criamos um dataframe temporário para treino
      train_df <- data.frame(y = dadosTy[, curr_var], dadosTx)
      
      # Metodo anova para regressão
      fit_tree <- rpart(y ~ ., data = train_df, method = "anova")
      
      exec_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
      
      # --- D. CÁLCULO DE MÉTRICAS ---
      # Para predição, o newdata deve ter as mesmas colunas do treino (V1, V2...)
      # Como dadosVx já tem essas colunas, podemos passar direto ou como dataframe
      ypred <- predict(fit_tree, newdata = data.frame(dadosVx))
      ypred_train <- predict(fit_tree)
      
      # Validação
      if (sd(ypred) == 0) { 
        r2_val <- 0 
      } else { 
        rv <- cor(dadosVy[, curr_var], ypred)
        r2_val <- as.numeric(rv^2)
      }
      rmse_val <- as.numeric(sqrt(mean((dadosVy[, curr_var] - ypred)^2)))
      
      # Treinamento
      rt <- cor(dadosTy[, curr_var], ypred_train)
      r2_train <- as.numeric(rt^2)
      rmse_train <- as.numeric(sqrt(mean((dadosTy[, curr_var] - ypred_train)^2)))
      
      # Salvar Métricas
      curr_metrics <- data.frame(
        Model = "DT", 
        Replicate = curr_rep, 
        Variable = curr_var, 
        Fold = fold_k, 
        R2_Train = r2_train, 
        R2_Val = r2_val, 
        RMSE_Train = rmse_train, 
        RMSE_Val = rmse_val, 
        Time_Sec = exec_time
      )
      
      write.csv(curr_metrics, sprintf("%s/METRICS_Rep%d_Var%d_Fold%d.csv", out_dt, curr_rep, curr_var, fold_k), row.names = FALSE)
      
      # --- E. IMPORTÂNCIA (Padrão Solicitado) ---
      
      imp <- tryCatch({ data.frame(varImp(fit_tree, value = "rss")) }, error=function(e) NULL)
      
      if (!is.null(imp)) {
         # --- PADRONIZAÇÃO DE COLUNAS (IMPORTÂNCIA) ---
         imp$Marker <- rownames(imp)
         colnames(imp)[1] <- "Overall"
         
         imp_final <- data.frame(
            Model = "DT",
            Replicate = curr_rep,
            Variable = curr_var,
            Fold = fold_k,
            Marker = imp$Marker,
            Overall = imp$Overall
         )
         
         fold_importance_accum <- rbind(fold_importance_accum, imp_final)
         write.csv(imp_final, outfile_imp, row.names = FALSE)
      } else {
         # Salva placeholder vazio
         dummy <- data.frame(Model="DT", Overall=0)
         write.csv(dummy, outfile_imp, row.names=FALSE)
      }
      
    }, error = function(e) {
       writeLines(paste("Erro no DT:", e$message), file.path(out_dt, "error_log.txt"))
    })
    
    # Limpeza de memória
    rm(fit_tree, train_df, dadosTx, dadosVx, dadosTy, dadosVy); gc()
  }
  return(NULL)
}

# 3. Consolidação Final
cat("Consolidando resultados do DT...\n")

metric_files <- list.files(out_dt, pattern = "METRICS_.*\\.csv", full.names = TRUE)
imp_files    <- list.files(out_dt, pattern = "IMP_.*\\.csv", full.names = TRUE)

if(length(metric_files) > 0) {
  dt_metrics <- map_dfr(metric_files, read.csv)
  write.csv(dt_metrics, paste0(out_dt, "/DT_Results_Metrics_CONSOLIDATED.csv"), row.names = FALSE)
}

if(length(imp_files) > 0) {
  dt_imp <- map_dfr(imp_files, read.csv)
  write.csv(dt_imp, paste0(out_dt, "/DT_Results_Importance_CONSOLIDATED.csv"), row.names = FALSE)
}

cat("DT finalizado.\n")
```

### 4.3. Bagging (BA)

Bagging is essentially a Random Forest where `mtry` (number of variables randomly sampled as candidates at each split) is equal to the total number of predictors (`ncol(X)`). This forces the trees to consider all markers at every split, reducing bias but potentially increasing correlation between trees compared to standard RF.

```{r ba_run, eval = FALSE}
# 1. Setup
grid_ba <- expand.grid(Rep = 1:n_reps, Variable = target_vars, stringsAsFactors = FALSE)
out_ba <-  paste0(base_out_dir,"/bagging")
if (!dir.exists(out_ba)) dir.create(out_ba, recursive = TRUE)

cat("Iniciando Bagging. Total de tarefas:", nrow(grid_ba) * nfolds, "\n")

set.seed(123)

# 2. Execução
ba_list <- foreach(task_i = 1:nrow(grid_ba), 
                   .packages = c('randomForest', 'caret', 'tidyverse'),
                   .export = c("filesFenT", "filesGenT", "filesFenV", "filesGenV", "nfolds")) %dopar% {
  
  curr_rep <- grid_ba$Rep[task_i]
  curr_var <- grid_ba$Variable[task_i]
  
  # Acumuladores locais
  fold_metrics_accum <- data.frame()
  fold_importance_accum <- data.frame()
  
  for (fold_k in 1:nfolds) {
    
    # A. Lógica de Reciclagem de Arquivos
    raw_idx <- (curr_rep - 1) * nfolds + fold_k
    file_idx <- ((raw_idx - 1) %% length(filesFenT)) + 1
    
    # Checkpoint
    outfile_imp <- sprintf("%s/IMP_Rep%d_Var%d_Fold%d.csv", out_ba, curr_rep, curr_var, fold_k)
    if(file.exists(outfile_imp)) next
    
    tryCatch({
      # B. Leitura dos Dados
      dadosTy <- read.table(filesFenT[file_idx], header = FALSE)
      dadosTx <- read.table(filesGenT[file_idx], header = FALSE)
      dadosVy <- read.table(filesFenV[file_idx], header = FALSE)
      dadosVx <- read.table(filesGenV[file_idx], header = FALSE)
      
      start_time <- Sys.time()
      
      # C. Modelo (Bagging = RF com mtry = total de colunas)
      # ntree = 500 é um padrão robusto
      bag_model <- randomForest(x = dadosTx, y = dadosTy[, curr_var], 
                                mtry = ncol(dadosTx), 
                                ntree = 500, 
                                importance = TRUE)
      
      exec_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
      
      # --- D. CÁLCULO DE MÉTRICAS ---
      ypred <- predict(bag_model, newdata = dadosVx)
      ypred_train <- predict(bag_model)
      
      # Validação
      if (sd(ypred) == 0) { 
        r2_val <- 0 
      } else { 
        rv <- cor(dadosVy[, curr_var], ypred)
        r2_val <- as.numeric(rv^2)
      }
      rmse_val <- as.numeric(sqrt(mean((dadosVy[, curr_var] - ypred)^2)))
      
      # Treinamento
      rt <- cor(dadosTy[, curr_var], ypred_train)
      r2_train <- as.numeric(rt^2)
      rmse_train <- as.numeric(sqrt(mean((dadosTy[, curr_var] - ypred_train)^2)))
      
      # Salvar Métricas
      curr_metrics <- data.frame(
        Model = "BA", 
        Replicate = curr_rep, 
        Variable = curr_var, 
        Fold = fold_k, 
        R2_Train = r2_train, 
        R2_Val = r2_val, 
        RMSE_Train = rmse_train, 
        RMSE_Val = rmse_val, 
        Time_Sec = exec_time
      )
      
      write.csv(curr_metrics, sprintf("%s/METRICS_Rep%d_Var%d_Fold%d.csv", out_ba, curr_rep, curr_var, fold_k), row.names = FALSE)
      
      # --- E. IMPORTÂNCIA (Padrão Solicitado) ---
      
      # Nota: Para RF/Bagging, varImp pode usar %IncMSE ou IncNodePurity.
      # O padrão do caret para RF é geralmente IncNodePurity (similar ao RSS em árvores).
      imp <- tryCatch({ data.frame(varImp(bag_model)) }, error=function(e) NULL)
      
      if (!is.null(imp)) {
         # --- PADRONIZAÇÃO DE COLUNAS (IMPORTÂNCIA) ---
         imp$Marker <- rownames(imp)
         colnames(imp)[1] <- "Overall"
         
         imp_final <- data.frame(
            Model = "BA",
            Replicate = curr_rep,
            Variable = curr_var,
            Fold = fold_k,
            Marker = imp$Marker,
            Overall = imp$Overall
         )
         
         fold_importance_accum <- rbind(fold_importance_accum, imp_final)
         write.csv(imp_final, outfile_imp, row.names = FALSE)
      } else {
         dummy <- data.frame(Model="BA", Overall=0)
         write.csv(dummy, outfile_imp, row.names=FALSE)
      }
      
    }, error = function(e) {
       writeLines(paste("Erro no Bagging:", e$message), file.path(out_ba, "error_log.txt"))
    })
    
    # Limpeza de memória
    rm(bag_model, dadosTx, dadosVx, dadosTy, dadosVy); gc()
  }
  return(NULL)
}

# 3. Consolidação Final
cat("Consolidando resultados do Bagging...\n")

metric_files <- list.files(out_ba, pattern = "METRICS_.*\\.csv", full.names = TRUE)
imp_files    <- list.files(out_ba, pattern = "IMP_.*\\.csv", full.names = TRUE)

if(length(metric_files) > 0) {
  ba_metrics <- map_dfr(metric_files, read.csv)
  write.csv(ba_metrics, paste0(out_ba, "/BA_Results_Metrics_CONSOLIDATED.csv"), row.names = FALSE)
}

if(length(imp_files) > 0) {
  ba_imp <- map_dfr(imp_files, read.csv)
  write.csv(ba_imp, paste0(out_ba, "/BA_Results_Importance_CONSOLIDATED.csv"), row.names = FALSE)
}

cat("Bagging finalizado.\n")
```

### 4.4. Random Forest (RF)

Unlike Bagging, Random Forest only considers a subset of features (`mtry`) at each split. For regression problems, the standard default is `mtry = p/3` (where `p` is the number of markers). This decorrelates the trees and often improves performance on high-dimensional genomic data.

```{r rf_run, eval = FALSE}
# 1. Setup
grid_rf <- expand.grid(Rep = 1:n_reps, Variable = target_vars, stringsAsFactors = FALSE)
out_rf <- paste0(base_out_dir,"/rf")
if (!dir.exists(out_rf)) dir.create(out_rf, recursive = TRUE)

cat("Iniciando RF. Total de tarefas:", nrow(grid_rf) * nfolds, "\n")

set.seed(123)

# 2. Execução
rf_list <- foreach(task_i = 1:nrow(grid_rf), 
                   .packages = c('randomForest', 'caret', 'tidyverse'),
                   .export = c("filesFenT", "filesGenT", "filesFenV", "filesGenV", "nfolds")) %dopar% {
  
  curr_rep <- grid_rf$Rep[task_i]
  curr_var <- grid_rf$Variable[task_i]
  
  # Acumuladores locais
  fold_metrics_accum <- data.frame()
  fold_importance_accum <- data.frame()
  
  for (fold_k in 1:nfolds) {
    
    # A. Lógica de Reciclagem de Arquivos
    raw_idx <- (curr_rep - 1) * nfolds + fold_k
    file_idx <- ((raw_idx - 1) %% length(filesFenT)) + 1
    
    # Checkpoint
    outfile_imp <- sprintf("%s/IMP_Rep%d_Var%d_Fold%d.csv", out_rf, curr_rep, curr_var, fold_k)
    if(file.exists(outfile_imp)) next
    
    tryCatch({
      # B. Leitura dos Dados
      dadosTy <- read.table(filesFenT[file_idx], header = FALSE)
      dadosTx <- read.table(filesGenT[file_idx], header = FALSE)
      dadosVy <- read.table(filesFenV[file_idx], header = FALSE)
      dadosVx <- read.table(filesGenV[file_idx], header = FALSE)
      
      start_time <- Sys.time()
      
      # C. Modelo RF Standard
      # mtry padrão para regressão é p/3
      rf_model <- randomForest(x = dadosTx, y = dadosTy[, curr_var], 
                               mtry = floor(ncol(dadosTx)/3), 
                               ntree = 500,
                               importance = TRUE)
      
      exec_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
      
      # --- D. CÁLCULO DE MÉTRICAS ---
      ypred <- predict(rf_model, newdata = dadosVx)
      ypred_train <- predict(rf_model)
      
      # Validação
      if (sd(ypred) == 0) { 
        r2_val <- 0 
      } else { 
        rv <- cor(dadosVy[, curr_var], ypred)
        r2_val <- as.numeric(rv^2)
      }
      rmse_val <- as.numeric(sqrt(mean((dadosVy[, curr_var] - ypred)^2)))
      
      # Treinamento
      rt <- cor(dadosTy[, curr_var], ypred_train)
      r2_train <- as.numeric(rt^2)
      rmse_train <- as.numeric(sqrt(mean((dadosTy[, curr_var] - ypred_train)^2)))
      
      # Salvar Métricas
      curr_metrics <- data.frame(
        Model = "RF", 
        Replicate = curr_rep, 
        Variable = curr_var, 
        Fold = fold_k, 
        R2_Train = r2_train, 
        R2_Val = r2_val, 
        RMSE_Train = rmse_train, 
        RMSE_Val = rmse_val, 
        Time_Sec = exec_time
      )
      
      write.csv(curr_metrics, sprintf("%s/METRICS_Rep%d_Var%d_Fold%d.csv", out_rf, curr_rep, curr_var, fold_k), row.names = FALSE)
      
      # --- E. IMPORTÂNCIA (Padrão Solicitado) ---
      
      imp <- tryCatch({ data.frame(varImp(rf_model)) }, error=function(e) NULL)
      
      if (!is.null(imp)) {
         # --- PADRONIZAÇÃO DE COLUNAS (IMPORTÂNCIA) ---
         imp$Marker <- rownames(imp)
         colnames(imp)[1] <- "Overall"
         
         imp_final <- data.frame(
            Model = "RF",
            Replicate = curr_rep,
            Variable = curr_var,
            Fold = fold_k,
            Marker = imp$Marker,
            Overall = imp$Overall
         )
         
         fold_importance_accum <- rbind(fold_importance_accum, imp_final)
         write.csv(imp_final, outfile_imp, row.names = FALSE)
      } else {
         dummy <- data.frame(Model="RF", Overall=0)
         write.csv(dummy, outfile_imp, row.names=FALSE)
      }
      
    }, error = function(e) {
       writeLines(paste("Erro no RF:", e$message), file.path(out_rf, "error_log.txt"))
    })
    
    # Limpeza de memória
    rm(rf_model, dadosTx, dadosVx, dadosTy, dadosVy); gc()
  }
  return(NULL)
}

# 3. Consolidação Final
cat("Consolidando resultados do RF...\n")

metric_files <- list.files(out_rf, pattern = "METRICS_.*\\.csv", full.names = TRUE)
imp_files    <- list.files(out_rf, pattern = "IMP_.*\\.csv", full.names = TRUE)

if(length(metric_files) > 0) {
  rf_metrics <- map_dfr(metric_files, read.csv)
  write.csv(rf_metrics, paste0(out_rf, "/RF_Results_Metrics_CONSOLIDATED.csv"), row.names = FALSE)
}

if(length(imp_files) > 0) {
  rf_imp <- map_dfr(imp_files, read.csv)
  write.csv(rf_imp, paste0(out_rf, "/RF_Results_Importance_CONSOLIDATED.csv"), row.names = FALSE)
}

cat("RF finalizado.\n")
```

### 4.5. Boosting (BO)

Gradient Boosting Machines (GBM) build an ensemble of weak prediction models (typically shallow trees) sequentially. Each new tree attempts to correct the errors (residuals) of the previous ones. Key parameters are `shrinkage` (learning rate) and `interaction.depth` (tree complexity).

```{r bo_run, eval = FALSE}
# 1. Setup
grid_bo <- expand.grid(Rep = 1:n_reps, Variable = target_vars, stringsAsFactors = FALSE)
out_bo <- paste0(base_out_dir,"/boosting")
if (!dir.exists(out_bo)) dir.create(out_bo, recursive = TRUE)

cat("Iniciando Boosting. Total de tarefas:", nrow(grid_bo) * nfolds, "\n")

set.seed(123)

# 2. Execução
bo_list <- foreach(task_i = 1:nrow(grid_bo), 
                   .packages = c('gbm', 'caret', 'tidyverse'),
                   .export = c("filesFenT", "filesGenT", "filesFenV", "filesGenV", "nfolds")) %dopar% {
  
  curr_rep <- grid_bo$Rep[task_i]
  curr_var <- grid_bo$Variable[task_i]
  
  # Local accumulators (optional since we save to disk)
  fold_metrics_accum <- data.frame()
  fold_importance_accum <- data.frame()
  
  for (fold_k in 1:nfolds) {
    
    # A. Lógica de Reciclagem de Arquivos
    raw_idx <- (curr_rep - 1) * nfolds + fold_k
    file_idx <- ((raw_idx - 1) %% length(filesFenT)) + 1
    
    # Checkpoint
    outfile_imp <- sprintf("%s/IMP_Rep%d_Var%d_Fold%d.csv", out_bo, curr_rep, curr_var, fold_k)
    if(file.exists(outfile_imp)) next
    
    tryCatch({
      # B. Leitura dos Dados
      dadosTy <- read.table(filesFenT[file_idx], header = FALSE)
      dadosTx <- read.table(filesGenT[file_idx], header = FALSE)
      dadosVy <- read.table(filesFenV[file_idx], header = FALSE)
      dadosVx <- read.table(filesGenV[file_idx], header = FALSE)
      
      start_time <- Sys.time()
      
      # C. Modelo
      # GBM exige dataframe e formula para funcionar bem com predict depois
      # Criamos um dataframe unificado para treino
      df_train <- data.frame(Y_TARGET = dadosTy[, curr_var], dadosTx)
      
      # Treinamento (Gaussian para traços contínuos)
      bo_model <- gbm(Y_TARGET ~ ., 
                      data = df_train, 
                      distribution = "gaussian",
                      n.trees = 500, 
                      interaction.depth = 2, 
                      shrinkage = 0.01,
                      verbose = FALSE)
      
      exec_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
      
      # --- D. CÁLCULO DE MÉTRICAS ---
      # Importante: gbm predict exige newdata com mesmas colunas do treino
      # dadosVx (Genótipo Validação) tem as mesmas colunas de dadosTx (V1, V2...)
      df_val <- data.frame(dadosVx)
      
      ypred <- predict(bo_model, newdata = df_val, n.trees = 500)
      ypred_train <- predict(bo_model, newdata = df_train, n.trees = 500)
      
      # Validação
      if (sd(ypred) == 0) { 
        r2_val <- 0 
      } else { 
        rv <- cor(dadosVy[, curr_var], ypred)
        r2_val <- as.numeric(rv^2)
      }
      rmse_val <- as.numeric(sqrt(mean((dadosVy[, curr_var] - ypred)^2)))
      
      # Treinamento
      rt <- cor(dadosTy[, curr_var], ypred_train)
      r2_train <- as.numeric(rt^2)
      rmse_train <- as.numeric(sqrt(mean((dadosTy[, curr_var] - ypred_train)^2)))
      
      # Salvar Métricas
      curr_metrics <- data.frame(
        Model = "BO", 
        Replicate = curr_rep, 
        Variable = curr_var, 
        Fold = fold_k, 
        R2_Train = r2_train, 
        R2_Val = r2_val, 
        RMSE_Train = rmse_train, 
        RMSE_Val = rmse_val, 
        Time_Sec = exec_time
      )
      
      write.csv(curr_metrics, sprintf("%s/METRICS_Rep%d_Var%d_Fold%d.csv", out_bo, curr_rep, curr_var, fold_k), row.names = FALSE)
      
      # --- E. IMPORTÂNCIA (Padrão Solicitado) ---
      # caret::varImp para gbm usa a redução de erro relativo
      imp <- tryCatch({ data.frame(varImp(bo_model, numTrees = 500)) }, error=function(e) NULL)
      
      if (!is.null(imp)) {
         # --- PADRONIZAÇÃO DE COLUNAS (IMPORTÂNCIA) ---
         imp$Marker <- rownames(imp)
         colnames(imp)[1] <- "Overall"
         
         imp_final <- data.frame(
            Model = "BO",
            Replicate = curr_rep,
            Variable = curr_var,
            Fold = fold_k,
            Marker = imp$Marker,
            Overall = imp$Overall
         )
         
         fold_importance_accum <- rbind(fold_importance_accum, imp_final)
         write.csv(imp_final, outfile_imp, row.names = FALSE)
      } else {
         dummy <- data.frame(Model="BO", Overall=0)
         write.csv(dummy, outfile_imp, row.names=FALSE)
      }
      
    }, error = function(e) {
       writeLines(paste("Erro no Boosting:", e$message), file.path(out_bo, "error_log.txt"))
    })
    
    # Limpeza de memória
    rm(bo_model, df_train, df_val, dadosTx, dadosVx, dadosTy, dadosVy); gc()
  }
  return(NULL)
}

# 3. Consolidação Final
cat("Consolidando resultados do Boosting...\n")

metric_files <- list.files(out_bo, pattern = "METRICS_.*\\.csv", full.names = TRUE)
imp_files    <- list.files(out_bo, pattern = "IMP_.*\\.csv", full.names = TRUE)

if(length(metric_files) > 0) {
  bo_metrics <- map_dfr(metric_files, read.csv)
  write.csv(bo_metrics, paste0(out_bo, "/BO_Results_Metrics_CONSOLIDATED.csv"), row.names = FALSE)
}

if(length(imp_files) > 0) {
  bo_imp <- map_dfr(imp_files, read.csv)
  write.csv(bo_imp, paste0(out_bo, "/BO_Results_Importance_CONSOLIDATED.csv"), row.names = FALSE)
}

cat("Boosting finalizado.\n")
```

## 5\. Result Consolidation

This step recursively searches for all individual result files generated by the Machine Learning models. It consolidates them into unified datasets, ensuring **strict data type consistency** and **cleaning Marker IDs** (removing prefixes like "V" or "M") to prevent merging errors.

```{r consolidate_ml}
cat("--- Starting Global ML Consolidation ---\n")

# Define root directory to search
ml_root_dir <- base_out_dir

if (dir.exists(ml_root_dir)) {
  
  # ==============================================================================
  # 1. CONSOLIDATE IMPORTANCE SCORES (For QTL Detection)
  # ==============================================================================
  
  # Recursive search for all Importance files (IMP_*.csv)
  imp_files <- list.files(path = ml_root_dir, pattern = "IMP_.*\\.csv", recursive = TRUE, full.names = TRUE)
  
  # Filter out any previously consolidated files
  imp_files <- imp_files[!grepl("CONSOLIDATED", imp_files)]
  
  if (length(imp_files) > 0) {
    cat(sprintf("Processing %d Importance files...\n", length(imp_files)))
    
    # Read and bind all files with TYPE SAFETY & CLEANING
    all_imp <- map_dfr(imp_files, function(f) {
      tryCatch({ 
        df <- read.csv(f)
        
        # --- LIMPEZA CRÍTICA DE MARCADORES ---
        if("Marker" %in% names(df)) {
          # 1. Forçar para texto (evita erro de integer vs character)
          df$Marker <- as.character(df$Marker)
          
          # 2. Remover letras/prefixos (Ex: "V100" -> "100", "M5" -> "5")
          # Mantém apenas os dígitos (0-9)
          df$Marker <- gsub("[^0-9]", "", df$Marker)
        }
        
        return(df)
      }, error = function(e) return(NULL))
    })
    
    # Standardize column names (Rep -> Replicate, Var -> Variable)
    if("Rep" %in% names(all_imp)) all_imp <- rename(all_imp, Replicate = Rep)
    if("Var" %in% names(all_imp)) all_imp <- rename(all_imp, Variable = Var)
    
    # Save Consolidated File (All Models)
    write.csv(all_imp, file.path(ml_root_dir, "ML_Results_Importance_CONSOLIDATED.csv"), row.names = FALSE)
    
    # Save Separate Files by Model
    if (nrow(all_imp) > 0) {
      unique_models <- unique(all_imp$Model)
      for (m in unique_models) {
        mod_data <- all_imp %>% filter(Model == m)
        out_name <- file.path(ml_root_dir, paste0(m, "_Results_Importance_CONSOLIDATED.csv"))
        write.csv(mod_data, out_name, row.names = FALSE)
        cat(sprintf("  -> Model %s Importance saved: %d rows.\n", m, nrow(mod_data)))
      }
    }
  } else {
    cat("No Importance files found.\n")
  }
  
  # ==============================================================================
  # 2. CONSOLIDATE PERFORMANCE METRICS (R2, RMSE)
  # ==============================================================================
  
  # Recursive search for all Metrics files (METRICS_*.csv)
  metric_files <- list.files(path = ml_root_dir, pattern = "METRICS_.*\\.csv", recursive = TRUE, full.names = TRUE)
  metric_files <- metric_files[!grepl("CONSOLIDATED", metric_files)]
  
  if (length(metric_files) > 0) {
    cat(sprintf("\nProcessing %d Metric files...\n", length(metric_files)))
    
    all_metrics <- map_dfr(metric_files, function(f) {
      tryCatch({ read.csv(f) }, error = function(e) return(NULL))
    })
    
    # Standardize columns
    if("Rep" %in% names(all_metrics)) all_metrics <- rename(all_metrics, Replicate = Rep)
    if("Var" %in% names(all_metrics)) all_metrics <- rename(all_metrics, Variable = Var)
    
    # Save Consolidated Metrics
    write.csv(all_metrics, file.path(ml_root_dir, "ML_Results_Metrics_CONSOLIDATED.csv"), row.names = FALSE)
    cat(sprintf("Global Metrics saved: %d rows.\n", nrow(all_metrics)))
    
  } else {
    cat("No Metric files found.\n")
  }
  
} else {
  cat("Output directory not found.\n")
}

cat("--- Consolidation Finished ---\n")
```

## 6\. Next Steps

With the Machine Learning results consolidated, the computational pipeline is complete.

The next step is the **[Consolidated Analysis](consolidated_analysis.html)**, where we will:

1.  Load the results from both GWAS and ML.
2.  Calculate **Detection Power**, **Precision**, and **F1 Scores**.
3.  Generate the final heatmaps comparing all methodologies.
