---
title: "ld decay"
author: "Costa, W. G."
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Importance-of-markers-for-QTL-detection-by-machine-learning-methods/
output:
  workflowr::wflow_html:
    toc: TRUE
  highlight: github
editor_options:
  chunk_output_type: console
github-repo: wevertongomescosta/Importance-of-markers-for-QTL-detection-by-machine-learning-methods
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries

```{r}
library(sommer)
library(ggplot2)
library(dplyr)
```

# Load genotype data

```{r}
CPgeno <- read.table("data/GEN.txt")
head(CPgeno[, 1:5])
```

# Load and prepare map data

```{r}
mapCP <- readRDS("data/map.rds") %>%
  mutate(
    Locus = paste0("V", 1:n()),
    Position = as.numeric(Tamanho),
    LG = as.integer(GL)
  ) %>%
  dplyr::select(Locus, Position, LG)

str(mapCP)
summary(mapCP)
```

# Run LD decay analysis for each linkage group

# Step 1: Obtain a sorted list of unique linkage groups from the map data

```{r}
unique_LGs <- sort(unique(mapCP$LG))
```

# Step 2: Apply a function to each linkage group using lapply
```{r}
res <- lapply(unique_LGs, function(lg) {
  # Step 2a: For the current linkage group 'lg', extract the corresponding marker names
  markers <- mapCP %>%
    filter(LG == lg) %>%  # Filter the map data to include only rows where LG equals the current linkage group
    pull(Locus)           # Extract the 'Locus' column, which contains marker names
  
  # Step 2b: Perform Linkage Disequilibrium (LD) decay analysis on the genotype data for these markers
  LDDecay <- LD.decay(CPgeno[, markers], # Subset the genotype matrix to include only the columns for the current markers
                      mapCP %>% filter(LG == lg))  # Subset the map data to include only the current linkage group
  
  # Step 2c: Filter the LD results to include only significant marker pairs after Bonferroni correction
  A <- LDDecay$all.LG %>%  # Access the LD results for all marker pairs
    filter(p < 0.05 / choose(length(markers), 2))  # Apply Bonferroni correction to adjust the p-value threshold
  
  # Step 2d: Check if there are any significant marker pairs
  if (nrow(A) > 0) {
    # If significant pairs exist, create a data frame including the linkage group identifier
    data.frame(GL = paste0("lg", lg), # Add a column 'GL' with the linkage group name (e.g., 'lg1', 'lg2', etc.)
               A)                       # Include the filtered LD decay results
  } else {
    # If no significant pairs are found, return NULL for this linkage group
    NULL
  }
})
```



# Combine results and remove NULLs
```{r}
res <- res[!sapply(res, is.null)]
dados <- do.call(rbind, res)
dados$GL <- as.factor(dados$GL)
```



# Fit the loess model
```{r}
mod <- loess(r2 ~ d, data = dados)
load("output/mod.rda")
save(mod, file = "output/mod.rda")
```

# Predict r² values for a sequence of distances
```{r}

d_seq <- seq(0, max(dados$d), 0.1)
pred <- predict(mod, d_seq)
data_pred <- data.frame(d = d_seq, pred = pred)
```

# Find the distance where r² is approximately 0.2
```{r}
target_r2 <- 0.87
closest_index <- which.min(abs(pred - target_r2))
closest_d <- d_seq[closest_index]
closest_d
```

# Plot the LD decay curve
```{r}
library(ggrepel)

# Create the plot with additional adjustments
color_palette <- colorRampPalette(c('black', 'gray50'))
p <- ggplot(dados, aes(x = d, y = r2, colour = GL)) +
  geom_point(pch = 21, show.legend = FALSE) +
  geom_line(
    data = data_pred,
    aes(x = d, y = pred),
    colour = 'red',
    linewidth = 1
  ) +
  geom_hline(yintercept = target_r2,
             linetype = "dashed",
             color = "blue") +
  geom_vline(xintercept = closest_d,
             linetype = "dashed",
             color = "green") +
  geom_point(
    aes(x = closest_d, y = target_r2),
    data = NULL,
    color = "gold",
    size = 3
  ) +
  scale_color_manual(values = color_palette(length(unique(dados$GL)))) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1)) +
  scale_x_continuous(breaks = seq(0, max(dados$d), by = 10)) +
  labs(x = "Distance (d)", y = expression(r^2)) +
  theme_classic() +
  theme(text = element_text(size = 20))

p +
  geom_text(
    aes(
      x = closest_d + 15,
      y = target_r2,
      label = paste("d =", round(closest_d, 2), "cM")
    ),
    data = NULL,
    nudge_y = 0.02,
    color = "black",
    fontface = "bold"
  )
```
