---
title: "GWAS - Multi-Model Parallel Analysis (GAPIT)"
author: "Costa, W. G."
date: "`r Sys.Date()`"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: TRUE
    toc_float: TRUE
    theme: flatly
    highlight: haddock
editor_options:
  chunk_output_type: console
---

# GWAS - High-Performance Multi-Model Analysis

This script performs a GWAS analysis utilizing multiple models (Single and Multi-Locus) within a cross-validation scheme or independent replicates.

**Objective**: To evaluate the stability of QTL detection.

**Models**: GLM, MLM, CMLM, MLMM, FarmCPU, Blink.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = "center"
)
```

## 1. Setup and Packages

This section establishes the computational environment required for the analysis. It performs a conditional check to install the **GAPIT** package directly from GitHub if it is not already present. Additionally, it loads essential libraries for data manipulation (`tidyverse`, `gtools`) and high-performance parallel computing (`doParallel`, `foreach`) to optimize the processing of multiple models and folds.

```{r packages, message=FALSE}
# Check and install GAPIT package from GitHub if not available
if (!require(GAPIT)) {
  devtools::install_github("jiabowang/GAPIT", force=TRUE)
  library(GAPIT)
}

# Install and load CRAN packages for data management and parallelism
if(!require(gtools)) install.packages("gtools")
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(doParallel)) install.packages("doParallel")
if(!require(foreach)) install.packages("foreach")

# Load libraries
library(gtools)     # For natural sorting of filenames
library(tidyverse)  # For data manipulation
library(doParallel) # For parallel backend
library(foreach)    # For parallel looping
library(GAPIT)      # For GWAS analysis
```

## 2. Data Preparation

In this step, we identify and list the partitioned files located in the `PHEN_T` (Phenotype) and `GEN_T` (Genotype) directories. We use natural sorting to ensure correct file ordering and perform integrity checks to guarantee that phenotype and genotype files are correctly paired. The genetic map is also loaded and standardized for GAPIT.

```{r load_data}
# 1. Define File Paths
pathFenT <- file.path("data/PHEN_T")
pathGenT <- file.path("data/GEN_T")
pathMap  <- "data/map.rds"

# 2. File Listing (Natural Numeric Sort)
# 'mixedsort' ensures that "file10" comes after "file2", rather than after "file1"
filesFenT <- gtools::mixedsort(list.files(path = pathFenT, pattern = "\\.txt$", full.names = TRUE))
filesGenT <- gtools::mixedsort(list.files(path = pathGenT, pattern = "\\.txt$", full.names = TRUE))

# Integrity Check
if(length(filesFenT) == 0 || length(filesGenT) == 0) {
  stop("CRITICAL ERROR: No phenotype or genotype files found in data/ folders.")
}

if(length(filesFenT) != length(filesGenT)) {
  warning("WARNING: Mismatch in number of phenotype and genotype files. Check pairing.")
}

# 3. Load Genetic Map
if(file.exists(pathMap)) {
  GM <- readRDS(pathMap) %>%
    # Forced standardization for GAPIT format
    rename(
      Name = any_of(c("marker", "SNP", "rs")),
      Chromosome = any_of(c("Chromosome", "Chr", "GL")),
      Position = any_of(c("Position", "Pos", "Tamanho"))
    ) %>%
    mutate(
      Name = as.character(Name),
      Chromosome = as.numeric(Chromosome),
      Position = as.numeric(Position)
    ) %>%
    select(Name, Chromosome, Position)
    
  cat("Genetic Map loaded:", nrow(GM), "markers.\n")
} else {
  stop("Map file 'data/map.rds' not found.")
}
```

## 3. Global Parameters and Cluster Setup

We define the core analysis parameters, including the number of cross-validation folds, independent replicates, target traits, and the specific GWAS models to be evaluated. Additionally, we configure the computing cluster for high-performance parallel execution.

```{r grid_setup}
# --- Experimental Parameters ---
nfolds <- 5
n_reps <- 10
total_traits <- 18
target_vars <- 1:total_traits 

# Model List (EXCLUDING SUPER)
models_list <- c("GLM", "MLM", "CMLM", "MLMM", "FarmCPU", "Blink", "SUPER")

# Output Directories
base_out_dir <- "output/gwas_cv"
temp_run_dir <- file.path(base_out_dir, "temp_runs") 
if (!dir.exists(temp_run_dir)) dir.create(temp_run_dir, recursive = TRUE)

# --- Cluster Configuration (High Performance) ---
# Without the SUPER model, we can maximize CPU usage
n_cores_total <- parallel::detectCores()
n_cores_use <- max(1, n_cores_total - 1)

cat("Cluster Configuration: Using", n_cores_use, "cores.\n")

if(exists("cl")) stopCluster(cl)
cl <- makeCluster(n_cores_use)
registerDoParallel(cl)
```

## 4. GWAS Execution (Cross-Validation)

Below, we iterate through all combinations of Replicate, Variable, and Model. Within each task, the 5 Folds are processed sequentially. Execution time is recorded, while the extraction of importance results is deferred to the consolidation stage.

```{r run_parallel, eval=FALSE}
# 1. Task Grid
task_grid <- expand.grid(
  Rep = 1:n_reps,
  Variable = target_vars,
  Model = models_list,
  stringsAsFactors = FALSE
)

cat(sprintf("Starting GWAS. Total Tasks: %d x %d Folds = %d analyses.\n", 
            nrow(task_grid), nfolds, nrow(task_grid)*nfolds))

# Global variables
main_wd <- getwd()

# 2. Parallel Loop
foreach(task_i = 1:nrow(task_grid), 
        .packages = c("GAPIT", "tidyverse"),
        .export = c("GM", "filesFenT", "filesGenT", "base_out_dir", "temp_run_dir", "main_wd", "nfolds")) %dopar% {
          
  curr_rep <- task_grid$Rep[task_i]
  curr_var <- task_grid$Variable[task_i]
  curr_model <- task_grid$Model[task_i]
  
  # Internal Loop over Folds
  for (fold_k in 1:nfolds) {
    
    # A. File Recycling Logic
    raw_idx <- (curr_rep - 1) * nfolds + fold_k
    file_idx <- ((raw_idx - 1) %% length(filesFenT)) + 1
    
    if (is.na(filesFenT[file_idx])) next 
    
    # B. Prepare Directory
    folder_name <- paste0("R", curr_rep, "_V", curr_var, "_M", curr_model, "_F", fold_k)
    task_dir <- file.path(main_wd, temp_run_dir, folder_name)
    
    # Check if already run
    if (file.exists(file.path(task_dir, "runtime_info.csv"))) next
    
    if (!dir.exists(task_dir)) dir.create(task_dir, recursive = TRUE)
    setwd(task_dir)
    
    tryCatch({
      # C. Data Loading
      pheno_path <- filesFenT[file_idx]
      geno_path  <- filesGenT[file_idx]
      
      if (!grepl("^(/|[A-Za-z]:|\\\\)", pheno_path)) {
         pheno_path <- file.path(main_wd, pheno_path)
         geno_path <- file.path(main_wd, geno_path)
      }

      pheno_t <- read.table(pheno_path, header = FALSE)
      geno_t  <- read.table(geno_path, header = FALSE)
      
      # Numeric Adjustment for GAPIT (0,1,2)
      if(min(geno_t, na.rm=T) < 0) { geno_t <- geno_t + 1 }
      
      taxa_ids <- paste0("Ind_", 1:nrow(geno_t))
      Y <- data.frame(Taxa = taxa_ids, Trait = pheno_t[, curr_var])
      GD <- data.frame(Taxa = taxa_ids, geno_t)
      
      # D. Run GAPIT (Optimized Standard Configuration)
      start_time <- Sys.time()
      
      capture.output({
        GAPIT(
          Y = Y, GD = GD, GM = GM,
          model = curr_model,
          PCA.total = 3,
          file.output = TRUE, 
          group.from = 1,        
          group.to = nrow(Y),      
          kinship.algorithm = "VanRaden",
          SNP.P3D = TRUE # Variance optimization (fast standard)
        )
      })
      
      end_time <- Sys.time()
      exec_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
      
      # E. Save Runtime
      write.csv(data.frame(
        Model = curr_model,
        Rep = curr_rep,
        Var = curr_var,
        Fold = fold_k,
        Time_Sec = exec_time
      ), "runtime_info.csv", row.names = FALSE)
      
    }, error = function(e) {
      writeLines(as.character(e), "error.log")
    })
    
    setwd(main_wd)
  }
  return(NULL)
}

stopCluster(cl)
cat("Parallel Analysis Finished.\n")
```

## 5. Result Consolidation

We use `map_dfr` logic to recursively read all raw results generated by GAPIT and consolidate them into a single structure. Metadata such as Replicate, Variable, Model, and Fold are extracted directly from the directory names. Special handling is applied to separate Manhattan plot data (NYC) from QQ plot data (Kansas) for multi-locus models like Blink and FarmCPU.

```{r consolidate_split}
cat("--- Starting Global Consolidation ---\n")

runs_dir <- file.path(base_out_dir, "temp_runs")

if (dir.exists(runs_dir)) {
  
  # Search for result files (GAPIT Pattern)
  files <- list.files(runs_dir, pattern = "GAPIT.Association.GWAS_Results.*\\.csv", recursive = TRUE, full.names = TRUE)
  
  if (length(files) > 0) {
    cat(sprintf("Processing %d files found...\n", length(files)))
    
    process_file <- function(f) {
      tryCatch({
        # 1. Parse Directory Metadata
        dir_name <- basename(dirname(f))
        parts <- strsplit(dir_name, "_")[[1]]
        
        rep_val  <- as.numeric(sub("^R", "", parts[1]))
        var_val  <- as.numeric(sub("^V", "", parts[2]))
        base_mod <- sub("^M", "", parts[3])
        fold_val <- as.numeric(sub("^F", "", parts[4]))
        
        # 2. Refine Model Name (NYC vs Kansas)
        file_name <- basename(f)
        final_model_name <- base_mod
        
        if (base_mod %in% c("Blink", "FarmCPU", "BLINK", "FARMCPU")) {
          if (grepl("NYC", file_name, ignore.case = TRUE)) {
            final_model_name <- paste0(base_mod, ".NYC")
          } else if (grepl("Kansas", file_name, ignore.case = TRUE)) {
            final_model_name <- paste0(base_mod, ".Kansas")
          }
        }
        
        # 3. Read and Standardize Data
        df <- read.csv(f, header = TRUE)
        if (nrow(df) == 0) return(NULL)
        
        # Identify P-value column robustly
        cols <- colnames(df)
        p_col_idx <- grep("^P\\.val", cols, ignore.case = TRUE)
        if (length(p_col_idx) > 0) colnames(df)[p_col_idx[1]] <- "P.value" else return(NULL)
        
        df %>%
          select(SNP, P.value) %>%
          mutate(
            Model = final_model_name,
            Replicate = rep_val,
            Variable = var_val,
            Fold = fold_val,
            Marker = as.character(SNP),
            Overall = -log10(pmax(P.value, 1e-300)) # Avoid log(0)
          ) %>%
          select(Model, Replicate, Variable, Fold, Marker, Overall)
          
      }, error = function(e) { return(NULL) })
    }
    
    # Execute processing
    all_results <- map_dfr(files, process_file)
    
    # Save Separate Files by Model (Recommended for memory management)
    if (nrow(all_results) > 0) {
        unique_models <- unique(all_results$Model)
        
        for (m in unique_models) {
          mod_data <- all_results %>% filter(Model == m)
          out_name <- file.path(base_out_dir, paste0(m, "_Results_Importance_CONSOLIDATED.csv"))
          write.csv(mod_data, out_name, row.names = FALSE)
          cat(sprintf("  -> Model %s saved: %d rows.\n", m, nrow(mod_data)))
        }
        
        # Save Global File (Optional, comment out if memory issues occur)
        write.csv(all_results, file.path(base_out_dir, "GWAS_Results_Importance_CONSOLIDATED.csv"), row.names = FALSE)
    }
    
  } else {
    cat("No result files found. Check if the parallel step completed successfully.\n")
  }
  
  # Consolidate Runtime Information
  time_files <- list.files(runs_dir, pattern = "runtime_info.csv", recursive = TRUE, full.names = TRUE)
  
  if (length(time_files) > 0) {
    
    # MODIFICAÇÃO: Usar uma função segura que checa se o arquivo tem conteúdo
    read_safe <- function(f) {
      # Só lê se o arquivo tiver mais de 0 bytes
      if (file.size(f) > 0) {
        tryCatch({
          return(read.csv(f))
        }, error = function(e) return(NULL))
      } else {
        return(NULL) # Pula arquivos vazios
      }
    }
    
    time_data <- map_dfr(time_files, read_safe)
    
    if (nrow(time_data) > 0) {
      write.csv(time_data, file.path(base_out_dir, "GWAS_Runtime_Summary.csv"), row.names = FALSE)
      cat("Runtime summary saved to GWAS_Runtime_Summary.csv\n")
    } else {
      cat("Warning: runtime_info files were found but contained no valid data.\n")
    }
  }
}
```

## 6. Conclusion and Next Steps

With the consolidated results saved in `output/gwas_cv`, we can now proceed to analyze the stability of QTL detection across different models and cross-validation folds.

The generated output files allow for:
1.  **Visualization:** Creation of Manhattan and QQ plots to assess signal strength and model inflation.
2.  **Statistical Comparison:** Evaluation of model consistency across replicates.
3.  **Candidate Selection:** Identification of robust marker associations.

The next phase of this workflow involves estimating marker importance using **[Genomic Prediction by Machine Learning methods](genomic_prediction.html)**.
